---
title: "Purrr::Map"
description: |
  This will provide a guide to using map() in R-studio.
author:
  - name: Shaina Trevino, Jonathan Rochelle, Andre Klest
    url: https://example.com/norajones
date: 06-12-2019
output:
  distill::distill_article:
    self_contained: false
---


```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(rio)
library(tidyverse)
library(broom)
library(glue)
```

## Importing data

```{r import, eval=FALSE}
drug_by_year <- import(here("data", "drug_by_year.csv"),
                       setclass = "tbl_df")

recent_drug <- import(here("data", "2015_drug_dep.csv"),
                      setclass = "tbl_df")

```





`purrr:map()` is a functional. It takes a vector (or list of vectors) and a function as it's input, runs the function for each element in the vector and then returns a list.

  * The `purrr` package is loaded when you load `tidyverse`
  
For example, suppose you have a list that contains two other numbered lists with 5 values rounded to two decimal places and you wanted to round them all to the nearest integer.

Our list looks like this:

```{r create-list}
#creating list from two vectors of decimal values
v <- c(1.85, 2.07, 3.71, 4.21, 5.36)
v1 <- c(1.23, 2.53, 3.44, 4.71, 5.96)
l <- list(v, v1)
l
```

If we only had 1 vector to round, we could use the `round()` function

```{r ex-round}
round(v)
```

However, this does not work when supplied our list

```{r error-code, error = TRUE}
round(l)
```

Instead, we need a function that will apply `round()` to every element in our list. That's where we can use `map()`

```{r map-round}
#takes vector and rounds up
map(l, round)
```

Notice how the output from `map()` is a list, and the `round()` function has been applied to each element in the list. 

You could also use `map()` with a single vector, however, in this case, the output is the same as when solely using the `round()` function. Additionally, you can adjust which decimal that you would like to round to by specifying the digits. 

```{r round-single-vec}
map(v, round, digits = 1)
```

However, as mentioned above, this output is a list. When you have an atomic vector as an input or a list of numbered lists of length 1, you can coerce your output to an atomic vector. This coercion allows the function to utilize values that work within the operation called on. 

#### `map()` variants
  
You can override the return of list and instead return an atomic vector by using variants of `map()`:

    * `map_dbl` returns vector of type double
    
    * `map_chr` returns a vector of type character
    
    * `map_lgl` returns a vector of type logical (e.g., `TRUE` or `FALSE`)
    
    * `map_int` returns a vector of type integer
    
For example, if we use `map_dbl` or `map_chr` for our previous example, we will have the same results, but the output will be different types. 

```{r map-ex}
#return vector of type double
map_dbl(v, round)
#return vector as character
map_chr(v, round)
```

Notice how the numbers are actually the same, but `map_dbl` returns a numeric vector, whereas `map_chr` returns a character vector.

What if we wanted to know if each element in our vector is of type double, we would want to return a logical vector that tells us if the value is double (e.g., `TRUE`) or not (e.g., `FALSE`)

```{r map-ex1}
#return vector of type logical
map_lgl(v, is.double)
```

This logical vector tells us that each element is of type double.

`map_int` returns an integer vector (i.e., whole numbers), therefore, we can use it to ask questions like: what is the length of each element in my vector?

```{r map-ex2}
#return vector of type integer
map_int(v, length)
```

In this example, there is only 1 value per element in the vector, so all the lengths are 1.

If we use the same code on our list object, it should tell us how many numbers are in each element of the list.

```{r map-ex2-b}
map_int(l, length)
```

There are 2 elements in our list and each has 5 values. 

### Information about the data

Data used in this tutorial came from the Center for Disease Control and Prevention's (CDC) National Health and Nutrition Examination Survey (NHANES). The study began in 1999 and the most recent data collected and publicly available are from 2015-2016.

These data and codebooks are available online [here](https://www.cdc.gov/nchs/nhanes/about_nhanes.htm).

NHANES surveys around 5,000 participants accross the U.S. each year. Surveys include demographic, socioeconomic, dietary, and health related questions, as well as medical records, physiological assessments, and laboratory test. 

For this tutorial we will use two data frames: 

1. `drug_by_year` includes data from 2005 to 2015 and reports on the people who responded yes to using illicit substances in the last year. 

2. `recent_drug` includes has data from 2015-2016 and shows the age of first drug use, a summed depression score, whether or not participants have been to rehab, have access to mental health services, and have health insurance. 

### Using `map()` in data frames

Since `map()` can take a vector for an input, and data frames are lists containing vectors, you can use `map` to iterate through data frames. 

For example, say we want to use the NHANES data to calculate the mean for age of first use for each drug and depression scores. Since there are quite a bit of missing values in this data set, so we will need to add arguments to the mean function to ignore the missing data. 

The data frame looks like this

```{r tidy-data-map-ex, echo = FALSE, eval=FALSE}
d <- recent_drug %>% 
  select(age_cocaine:sum_depression)
d
```

To add arguments to a function within `map()`, we need to create an inline anonymous function where your function is first specified with `function(x)` where `x` is a place holder for what the function is going to run on. Then you write a function that you want to run on each element (`x`). 

Therefore, to calculate the means, our `map()` function would look like this

```{r summarize, eval=FALSE}
means <- map_dbl(d, function(x) mean(x, na.rm = TRUE)) 
means
```

There is also a shortcut you can use within `map()` functions where you don't have to specify `function(x)`. 

```{r summarize1, eval=FALSE}
means <- map_dbl(d, ~ mean(.x, na.rm = TRUE))
means
```

These two examples are exactly the same, the latter just has less code. `function(x)` is implied with the `~` and then you just have to use `.x` instead of `x` for what you are iterating over.

#### `map_df()`

There is another `map()` variant that is useful when working with data frames, `map_df()`.

You can use `map_df()` to return a data frame as the output instead of an atomic vector. 

For example, say we want to return a data frame of the means and standard deviations of each column within our data frame. 

You would expect the below example to work, but it doesn't because only the last line in the function is being evaluated and printed, so we only get standard deviations as the output. Also note you need to use curly braces for longer functions as seen here.

```{r map-df, eval=FALSE}
map_df(d, function(x) {
  mean(x, na.rm = TRUE)
  sd(x, na.rm = TRUE)
}) 
```

Instead, you need to instead use `summarize()` to work within the data frame and `summarize()` each vector

```{r map-df1, eval=FALSE}
#creating a vector for the column names to add to data frame
column_names <- colnames(d) 

map_df(d, function(x) {
  summarize(d, mean = mean(x, na.rm = TRUE),
            sd = sd(x, na.rm = TRUE))
}, .id = "column_names")
```

Or you could do the above in two separate steps, using `map()` to loop the functions through vectors and then put the vectors into a data frame

```{r map-df2, eval=FALSE}
#first create vectors by looping functions through data frames
mean <- map(d, ~mean(.x, na.rm = TRUE))
sd <- map(d, ~sd(.x, na.rm = TRUE))
#then combine vectors
cv <- c(mean, sd, recursive = TRUE)
#manually change attributes
attr(cv, "dim") <- c(6, 2) #6 variables in two columns (mean and sd)
colnames(cv) <- c("mean", "sd")
#transform to data frame
cv <- data.frame(cv, row.names = column_names)
cv
```

Parallel Iteration (e.g., `map2_*`, `pmap_*`) 

The `map` variants discussed above all take a single input to iterate (i.e., call the function on), but sometimes you need to iterate through more than just one input. 

This can be accomplished with `map2_` or `pmap_`

* `map2_` iterates over two inputs (e.g., two vectors)

* `pmap_` iterates over any number of inputs

For example, say you wanted to calculate a weighted mean from a vector of data values and a vector of weights.

```{r parallel, eval=FALSE}
#create vectors for simulated data from mean distribution of our data and weights
obs <- map(1:1715, ~rnorm(n = 20, mean = 17.289, sd = 4.558))
wts <- map(1:1715, ~rpois(20, 5))
```

The following example doesnt work because the weight argument is expecting a single value to apply to all observations, but we have a vector of weights. 

```{r parallel1, error = TRUE, eval=FALSE}
map_dbl(obs, ~weighted.mean(.x, w = wts))
```

Instead, we need a parallel iteration with `map2()` to run the function on each vector of observation with each corresponding vector of weights. 

Also, notice that when using `map2()` you will have an additional argument `.y`, along with `.x`, that corresponds to the second vector you are iterating through. 

```{r parallel2, eval=FALSE}
#note that we used head() to prevent the entirety of the extensive list to be displayed
head(map2_dbl(obs, wts, ~weighted.mean(.x, .y)))
```

The results is a weighted mean score for each participant. 

#### `pmap()` briefly

`pmap()` can be supplied any number of inputs that you provide in a single list. This list can contain any number of inputs or arguments. 

`pmap()` is very useful when working with data frames in which there are many vectors of the same length that can be provided as inputs (i.e., columns)

You can name the inputs in the list which is helpful when iterating through complex functions

We will show you how to use `pmap` to iterate and build custom plots in a later blog post


#### Other `purrr` functions (e.g., `walk_*`, `reduce`, `modify_*`) 

The `map()` variants discussed above all return an output (either an atomic vector or list). However, sometimes you don't need to return an atomic vector or list, but iteration would still be helpful.
  
Very useful variant of `map()` is `walk()`, specifically `walk2()` which can be used to save objects to a disk. Note: when saving something to a disk you have to provide 2 inputs: the object and the file path - this requires parallel iteration.   

The following code splits our longitudinal dataset by year and saves a separate .csv file for each year with the title ("year-2005", etc.) in the data folder of the working directory

```{r walk-ex, eval = FALSE}
by_year <- split(drug_by_year, drug_by_year$year)
paths <- here("data", paste0("year-", names(by_year), ".csv"))
walk2(by_year, paths, write.csv)


list.files("data")
```


Other useful functions are: 

* `reduce()` takes a vector of any length as it's input and returns a vector of length 1. For example `reduce(1:4, f)` returns `f(f(f(1, 2), 3), 4)`. 

* `modify()` returns the same type of input it is provided. For example, `modify(df, f)` will run the function on each column of the data frame and return a data frame as the output.

#### Using `mutate` within `nest()` is powerful in list columns

The data frames we have worked with so far have all contained columns of atomic vectors, however you can also have columns of lists within data frames (i.e., list column).

You can use `group_by()` and `nest()` to split your data into list columns then apply `mutate` to:

1. fit models by the grouping factor and extract coefficients

2. compare model fits

3. extract means from models fit with two groups

4. fit multiple different models for each grouping factor and view coefficients


First, we need to transform our data a bit so that we can properly fit our models.

```{r nest-mutate1-clean, eval=FALSE}
dat <- recent_drug %>% 
  select(-1:-2, -age_stim) %>%
  gather(drug, age_use, -age, -sum_depression:-sex) %>%
  separate(drug, c("dis", "drug"), sep = "_") %>%
  select(-dis) %>% 
  filter(!is.na(age_use)) %>% 
  mutate(mental_health = ifelse(mental_health == "Yes", "Yes", "No"),
         sex = fct_relevel(sex, ref = "Male"),
         mental_health = fct_relevel(mental_health, ref = "No"))
```

Then, we will use `group_by()` and `nest()` to split our data by each type of drug, fit a model that predicts age of first use from participant sex for each drug category, and view the coefficients.

```{r nest-mutate1, eval=FALSE}
#fit one model per drug and view coefficients
by_drug_coefs <- dat %>% 
  group_by(drug) %>% 
  nest() %>% 
  mutate(model = map(data, ~lm(age_use ~ sex, data = .x)),
         coef = map(model, ~tidy(.x))) %>% 
  unnest(coef)
by_drug_coefs
```

Next, we can compare all model fits (note the model fits will not be good since our models were not significant).

```{r nest-mutate2, eval=FALSE}
#Look at model fit stats per drug model
by_drug_fit <- dat %>% 
  group_by(drug) %>% 
  nest() %>% 
  mutate(model = map(data, ~lm(age_use ~ sex, data = .x)),
         fit = map(model, ~glance(.x))) %>% 
  unnest(fit)
by_drug_fit
```

Finally, we can also extract the mean age of first use for males and females for each drug

```{r nest-mutate3, eval=FALSE}
#mean age of use for males and females in first model
by_drug_sex <- dat %>% 
  group_by(drug) %>% 
  nest() %>% 
  mutate(model = map(data, ~lm(age_use ~ sex, data = .x)),
         coefs = map(model, coef),
         males = map_dbl(coefs, 1),
         females = males + map_dbl(coefs, 2)) %>% 
  select(drug, males, females)
by_drug_sex
```

Another powerful use of list columns is being able to fit multiple models by each drug. Then, you could run through the same steps as above to compare model fits, etc. 

``` {r nest-mutate4, eval=FALSE}
#fit multiple models by year
by_drug_models <- dat %>% 
  group_by(drug) %>% 
  nest() %>% 
  mutate(m1 = map(data, ~lm(age_use ~ sex, data = .x)),
         m2 = map(data, ~lm(age_use ~ sex + mental_health, data = .x)),
         m3 = map(data, ~lm(age_use ~ sex * mental_health, data = .x)))
```


```{r nest-mutate, include = FALSE, eval=FALSE}

d_long <- drug_by_year %>% 
  select(-V1, -ethnicity) %>% 
  mutate(year = factor(year),
         drug = factor(drug),
         sex = factor(sex),
         response = factor(response))

by_year_drug <- d_long %>% 
  group_by(year, drug) %>% 
  nest() %>% 
  mutate(count = map(data, ~mutate(.x, count = n()))) 
```
